---
title: 'Class 11: Alpha Fold Pt. 2'
author: "Brittney Hayes"
date: "2024-02-13"
output: word_document
---

Alphafold was run via GoogleColab at https://colab.research.google.com/github/sokrypton/ColabFold

We used AplhaFold2_mmseqs2 version.

The main outputs include a set of **PDB structure files** along with matching **JSON format files** that tell us how good the resulting models might be.

## Custom analysis of resulting models

Let's start by loading the PDB structures in Mol*

```{r}
library(bio3d)

results_dir <-"HIVPrDimer_23119"

# File names for all PDB models
list.files(path ="HIVPrDimer_23119", pattern="*.pdb",
                        full.names = TRUE)
pdb_files <- list.files(path= "HIVPrDimer_23119",
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

# Quick view of model sequences

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```


```{r}
pdbs
```

# RMSD

```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

# Draw a heatmap of these RMSD matrix values

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

# Plot the pLDDT values across all models

```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

# Improve the superposition/fitting of our models by finding the most consistent “rigid core”

```{r}
core <- core.find(pdbs)
```
# Use the identified core atom positions as a basis for a more suitable superposition and write out the fitted structures

```{r}
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

# Examine the RMSF between positions of the structure

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

## Predicted Alignment Error for domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

# For example purposes lets read the 1st and 5th files

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

# Checking other PAE values
```{r}
pae1$max_pae
```

```{r}
pae5$max_pae
```

#Plot the N by N (where N is the number of residues) PAE scores with ggplot or with functions from the Bio3D package

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

# Plot all of these using the same z range

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

## Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

# How many sequences are in this alignment

```{r}
dim(aln$ali)
```

We can score residue conservation in the alignment with the conserv() function.

```{r}
sim <- conserv(aln)

plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

# Generate a consensus sequence with a high cutoff value

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

# Map this conservation score to the Occupancy column of a PDB file for viewing in molecular viewer programs

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```

